Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-04-19 manuscript/main.qmd
upload r squared results and methods changes

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF). We then randomly sampled (1000 iterations) these gradients using a decreasing sample size (50, 20, 10, and 5) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**). Subsequently, we test whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples, providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.

To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?. To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples. 5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

>>>>

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF). We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**). Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.

we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).
